The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-20 paper/ms.tex
Fix previous commit
Update text to change to 10% error as neighborhood

\authors{others}

@@

\subsection{The size of the solar neighborhood} \label{sssec:neighborhood}
Given a local coordinate system, the amplitude and characteristic length scale
of the density fluctuations in the Galactic disk will determine how far this
system can be extended before introducing $\mathcal{O}(1)$ systematic errors
in the computation of actions for thin-disk stars. We consider this a
criterion for the region around the Sun where the analysis of the action-space
distribution of stars on quasi-circular orbits, and therefore the epicyclic
approximation, can be expected to be valid. This leads to a somewhat
dynamically motivated definition of the ``solar neighborhood.'' Since the
motivation may be different for defining the solar neighborhood in, e.g.,
studies of the chemical distribution of the Galaxy, we propose the phrase
``dynamical solar neighborhood.''

@@

\begin{figure*}
\plotone{fig/range_dphi.pdf}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $100\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

From Figure~\ref{fig:range_deltaphi} it is clear that the range of midplane
values spanned in a given $\Delta \phi$ quickly saturates at $\Delta \phi
\gtrsim \pi/2$ radians (i.e. one quarter of the way around the Galactic disk
from the Sun). At far smaller $\Delta \phi$ it reaches the level of deviation
required for an $\mathcal{O}(1)$ systematic error in $J_z$ for our fiducial
thin-disk-like orbit (\thincolor{} horizontal line in Figure
\ref{fig:range_deltaphi}). For the three simulations considered, the median
range exceeds this value at $\Delta \phi/\pi > 0.072, 0.11,
0.34\,\text{rad}$, with corresponding physical scales of $\lambda \sim 900,
1400, 4300\,\pc$ for \mi{}, \mf{}, and \mm{}, respectively. These can be
considered estimates of the range of validity for an extension of the locally
determined axisymmetric coordinate system, at least when it comes to the study
of stars on quasi-circular orbits.

The situation is more forgiving for the study of thick-disk-like orbits, where
an $\mathcal{O}(1)$ systematic error in $J_z$ requires a larger midplane
deviation (\thickcolor{} horizontal line in Figure~\ref{fig:range_deltaphi}).
In fact, the range of midplane variation in \mi{} just barely saturates to
cause such an error in $J_z$, with the midplane variation in \mf{} and \mm{}
never reaching such a value. For stars on halo-like orbits, extending the
local coordinate system globally is not likely to cause serious problems,
since the midplane deviations are relatively small compared to those needed to
produce $\mathcal{O}(1)$ systematic errors in the actions (\halocolor{}
horizontal line in Figure \ref{fig:range_deltaphi}). However, we caution that
the tendency of these midplane deviations is to \emph{inflate} the value of
$J_z$ whenever the midplane deviation is $\gtrsim z_{\text{max}}$, scattering
them into the region of action space normally associated with different
stellar populations and possibly contributing to confusion there (see
Figure~\ref{fig:Jz_hist} and discussion therein). This inflation is extremely
non-Gaussian (as the induced error is for all orbits, e.g.
Figure~\ref{fig:Jz_hist}), and is particularly troubling for the case of thin
disk stars, where the necessary midplane deviation is
$\sim50\textup{--}100\,\pc$.

@@

Figure~\ref{fig:Rn_mc} shows that azimuthal midplane fluctuations will limit
the volume over which actions can be used to match stars to open clusters on
thin-disk-like orbits (\thincolor), depending on their mass. For
thick-disk-like orbits (\thickcolor), the tolerances are much more forgiving:
stars' actions can be compared over the entire Solar circle for nearly all
cluster masses, except at the very low-mass end in \mi . The neighborhood
around $\sim100\Msun$ clusters can be as small as $\sim300\,\pc$ but up to
$\sim1.2\,\kpc$ for \mi{} and \mf{}, though the situation is marginally better
for \mm{} with values ranging from $\sim0.8\,\kpc$ to $\sim3\,\kpc$. The
neighborhood values steadily increase with $m_c$, topping out at $\sim5\,\kpc$
for $\sim10^4\,\Msun$ clusters in
\mi{} and \mf{}. In \mm{}, the neighborhood size reaches the maximum allowed of
$16.4\,\kpc$ for clusters of mass $\sim800\,\Msun$. This figure also
underlines that the degree to which such an analysis will succeed depends
crucially on the expected degree of midplane fluctuations in the real Galaxy.
If the Milky Way is more like \mi , the range of action-space analysis
techniques like this is far more limited than if it more closely resembles
\mm{}.

We also compute $R_{\n}$ for each of the orbits of the real open clusters in
Table~\ref{tab:real_clusters}, using the midplane variations from
\mi{} as a conservative choice, since it has the largest degree of
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neigbhorhood around each cluster is typically $\gtrsim 500\,\pc$. This
analysis assumes that the MW resembles \mi{}, which is not necessarily true.
However, this step of our analysis is meant to be more illustrative than
quantitative: this calculation should actually be performed using real
measurements of variations in the MW's stellar midplane. Such measurements
should be possible in the near future with {\em Gaia}, with quality distance
estimates the crucial limiting factor.

@@

\item We interpret this time evolution as an error in the computed action
(Figure~\ref{fig:many_orbit_wrong_ref}). We focused on the effect that an
inaccuracy in the midplane has on the vertical action $J_z$. We found that in
order for there to be a $100\%$ error in $J_z$ (as defined by the middle
$90\uth$ percentile) one needs a midplane error of $\sim30\,\pc$ and
$\sim225\,\pc$ for a typical thin and thick disk orbit, respectively. The
fractional error is significantly less for halo orbits.

>>>>

\author{others}

@@

\subsection{The size of the solar neighborhood} \label{sssec:neighborhood}
Given a local coordinate system, the amplitude and characteristic length scale
of the density fluctuations in the Galactic disk will determine how far this
coordinate system can be extended before introducing relevant systematic errors
in the computation of actions for thin-disk stars. We consider a $10\%$ error to be a
criterion for the region around the Sun where the analysis of the action-space
distribution of stars on quasi-circular orbits, and therefore the epicyclic
approximation, can be expected to be valid. This leads to a somewhat
dynamically motivated definition of the ``solar neighborhood.'' Since the
motivation may be different for defining the solar neighborhood in, e.g.,
studies of the chemical distribution of the Galaxy, we propose the phrase
``dynamical solar neighborhood.'' We also point out that our use of a $10\%$ error is arbitrary.

@@

\begin{figure*}
\plotone{fig/range_dphi.pdf}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $10\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

From Figure~\ref{fig:range_deltaphi} it is clear that the range of midplane
values spanned in a given $\Delta \phi$ quickly saturates at $\Delta \phi
\gtrsim \pi/2$ radians (i.e. one quarter of the way around the Galactic disk
from the Sun). At far smaller $\Delta \phi$ it reaches the level of deviation
required for a $10\%$ systematic error in $J_z$ for our fiducial
thin-disk-like orbit (\thincolor{} horizontal line in Figure
\ref{fig:range_deltaphi}). For the three simulations considered, the median
range exceeds this value at $\Delta \phi/\pi > x, y,
z\,\text{rad}$, with corresponding physical scales of $\lambda \sim a,
b, c\,\pc$ for \mi{}, \mf{}, and \mm{}, respectively. These can be
considered estimates of the range of validity for an extension of the locally
determined axisymmetric coordinate system, at least when it comes to the study
of stars on quasi-circular orbits.

The situation is more forgiving for the study of thick-disk-like orbits, where
a $10\%$ systematic error in $J_z$ requires a larger midplane
deviation (\thickcolor{} horizontal line in Figure~\ref{fig:range_deltaphi}).
In fact, the range of midplane variation in \mi{} just barely saturates to
cause such an error in $J_z$, with the midplane variation in \mf{} and \mm{}
never reaching such a value. For stars on halo-like orbits, extending the
local coordinate system globally is not likely to cause serious problems,
since the midplane deviations are relatively small compared to those needed to
produce $10\%$ systematic errors in the actions (\halocolor{}
horizontal line in Figure \ref{fig:range_deltaphi}). However, we caution that
the tendency of these midplane deviations is to \emph{inflate} the value of
$J_z$ whenever the midplane deviation is $\gtrsim z_{\text{max}}$, scattering
them into the region of action space normally associated with different
stellar populations and possibly contributing to confusion there (see
Figure~\ref{fig:Jz_hist} and discussion therein). This inflation is extremely
non-Gaussian (as the induced error is for all orbits, e.g.
Figure~\ref{fig:Jz_hist}), and is particularly troubling for the case of thin
disk stars, where the necessary midplane deviation is
$\sim5\textup{--}10\,\pc$.

@@

Figure~\ref{fig:Rn_mc} shows that azimuthal midplane fluctuations will limit
the volume over which actions can be used to match stars to open clusters on
thin-disk-like orbits (\thincolor), depending on their mass. For
thick-disk-like orbits (\thickcolor), the tolerances are much more forgiving:
stars' actions can be compared over the entire Solar circle for nearly all
cluster masses, except at the very low-mass end in \mi . The neighborhood
around $\sim100\Msun$ clusters can be as small as $\sim x\,\pc$ but up to
$\sim y\,\kpc$ for \mi{} and \mf{}, though the situation is marginally better
for \mm{} with values ranging from $\sim a\,\kpc$ to $\sim b\,\kpc$. The
neighborhood values steadily increase with $m_c$, topping out at $\sim c\,\kpc$
for $\sim10^4\,\Msun$ clusters in
\mi{} and \mf{}. In \mm{}, the neighborhood size reaches the maximum allowed of
$16.4\,\kpc$ for clusters of mass $\sim m\,\Msun$. This figure also
underlines that the degree to which such an analysis will succeed depends
crucially on the expected degree of midplane fluctuations in the real Galaxy.
If the Milky Way is more like \mi , the range of action-space analysis
techniques like this is far more limited than if it more closely resembles
\mm{}.

We also compute $R_{\n}$ for each of the orbits of the real open clusters in
Table~\ref{tab:real_clusters}, using the midplane variations from
\mi{} as a conservative choice, since it has the largest degree of
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neigbhorhood around each cluster is typically $\gtrsim x\,\pc$. This
analysis assumes that the MW resembles \mi{}, which is not necessarily true.
However, this step of our analysis is meant to be more illustrative than
quantitative: this calculation should actually be performed using real
measurements of variations in the MW's stellar midplane. Such measurements
should be possible in the near future with {\em Gaia}, with quality distance
estimates the crucial limiting factor, along with careful selction function modelling.

@@

\item We interpret this time evolution as an error in the computed action
(Figure~\ref{fig:many_orbit_wrong_ref}). We focused on the effect that an
inaccuracy in the midplane has on the vertical action $J_z$. We found that in
order for there to be a $10\%$ error in $J_z$ (as defined by one half the middle
$90\uth$ percentile) one needs a midplane error of $\sim6\,\pc$ and
$\sim50\,\pc$ for a typical thin and thick disk orbit, respectively. The
fractional error is significantly less for halo orbits.
